---
title: "Exercise Performance Prediction"
author: "Nel Antuña"
date: "18/12/2020"
output: html_document
---

# Introduction

The purpose of this project will be to predict the way in which certain people perform some training exercises by analysing information gathered from body devices. 

Due to computer limitations, the type of prediction made will be pretty simple so that my laptop can run it. 

# Initial options

For this project we will mostly work with the following packages:

- Tidyverse package -> for data manipulation
- Caret package -> for machine learning operations

In addition, we will set seeds so that results can be replicable. 

```{r options, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
set.seed(255)
```

# Data exploration

## Data source

Data has been obtained from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har

```{r data source}
setwd("C:/Users/Nel/Documents/R files/college_major_analysis/")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "training.csv")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "testing.csv")
dataset <- read.csv("./training.csv")
```

## Data exploration

```{r data exploration, echo=FALSE}
dims <- dim(dataset)
```

The dataset downloaded has 'r dims[2]` variables and `r dims[1]` observations. 

```{r data exploration2, echo=FALSE}
table(sapply(dataset, class))
```

## Covariate selection

We will examine the dataset and select those variables relevant for the analysis

### Non-relevant variables

There are a series of variables that have no relation to the activity performance:

- Variable 'X' -> only gives the row number
- Time variables -> not required for the prediction
- User name -> not needed
- New and num window -> not relevant

```{r non relevant, echo=TRUE}
time_var <- colnames(select(dataset, contains("time")))
non_relevant <- c("X",time_var,"user_name","new_window","num_window")
```

### Detecting variables with high amount of NAs

We label those variables with over 90% of missing values

```{r nas, echo=TRUE}
nas <- rep(FALSE, ncol(dataset))
for (i in 3:ncol(dataset)) {
 t <- tapply(dataset[,i],dataset$user_name, function(x) mean(is.na(x)))
 if (mean(t) > 0.90) {nas[i] <- TRUE}
}
cols_with_nas <- colnames(dataset)[nas]
```

### Detecting variables with high amount of incomplete data

We label those variables with over 90% of incomplete data

```{r incomplete, echo=TRUE}
nulls <- apply(dataset, 2, function(x) mean(x == ""))>0.90
cols_with_nulls <- colnames(dataset)[nulls]
cols_with_nulls <- cols_with_nulls[!is.na(cols_with_nulls)]
```

## Final dataset

```{r final dataset, echo=TRUE, message=FALSE} 
non_var <- c(non_relevant,cols_with_nas,cols_with_nulls)
dataset_new <- select(dataset, -non_var)
dims_new <- dim(dataset_new)
```

The new dataset contains same number of observations (`r dims_new[1]`) but only `r dims_new[2]` variables

### Outcome

Outcome is Classe variable: category variable with the following options

```{r outcome, echo=FALSE} 
table(dataset_new$classe)
```

### Predictors

Below is displayed the number of predictors finally consider and their class:

```{r predictors, echo=FALSE} 
table(sapply(dataset_new[,-53],class))

```

A high-level exploration of the data shows that predictors don't seem to adhere to the normal distributions and correlation among them might lead to problems. Below an example with 5 of the variables:

```{r predictors fig, echo=FALSE, message=FALSE, warning=FALSE} 
featurePlot(x = dataset_new[,1:5], y = dataset_new$classe, plot = "pairs")
```

# Prediction analysis

## Setting up training and testing data sets

We start by creating a partition of the data for training and testing datasets

```{r partition, echo=TRUE, message=FALSE, warning=FALSE} 
inTrain <- createDataPartition(y = dataset_new$classe, p=0.8, list=FALSE)
training <- dataset_new[inTrain,]
testing <- dataset_new[-inTrain,]
```

## Training a model

We will use the training data set to produce our models. 
Characteristics:

- Algorithm will be classification tree
- Given the high amount of variables, we will include a Pricipal Components Analysis in order to reduce the impact that high number of variables can have on results 
- Given that variables don't adjust to normal distribution, a BoxCox correction will be inlcuded 

```{r model, message=FALSE, warning=FALSE} 
fit_ctrees <- train(classe~., data = training, method = c("ctree", "PCA", "BoxCox"))

```


## Accuracy 

```{r acc, message=FALSE, warning=FALSE} 
pred_ctrees_test <- predict(fit_ctrees, testing)

acc_ctrees <- (sum(pred_ctrees_test==testing$classe)/length(pred_ctrees_test))
```

When compare the prediciton of the testing values with the actual values, we see that the accuracy of the model is of `r acc_ctrees` 

# Prediciton

The initial dataset included 20 observations with ououtcome assign. 
We will predict the outcome of this data:

First, we need to prepare the data:
```{r test data, message=FALSE, warning=FALSE} 
test_data <- read.csv("./testing.csv")
#non_var object contained those variables that are not useful or have missing/incomplete values
test_data_new <- select(test_data, -non_var)
```

We run prediction on the data and these are the results:

```{r test data2, echo=FALSE, message=FALSE, warning=FALSE} 
predict(fit_ctrees, test_data_new)
```
